Beyond May: Complexity-stability relationships \\ in disordered dynamical systems

### 2023-05-02 papers/equilibrium.tex
update draft

Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=-x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
(Any coefficients can be reabsorbed in the statistics of $A$ and by a rescaling of time. 
We factor out a minus sign in order to connect more directly with competition models.)

\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{figs/cavity.pdf}
    \caption{The equilibrium distribution with power-law self- and cross-interactions is accurately reproduced by 
    Eq.\eqref{eq: dist general} and its Gaussian approximation.
    For simulations (in green) we used $\alpha=1$, $\beta=1.5$,
    $\gamma=1$, $N=50$, $\mu=\mu_s=10^{-2}$, $\sigma=5\cdot 10^{-3}$.}
    \label{fig: cavity sol.}
\end{figure}

Following e.g. Ref.~\cite{Roy2019}, we can derive from the evolution equation \eqref{dynamics} a dynamical mean field theory (DMFT) 
which describes the ensemble of $N\gg 1$ variables
by means of a single, self-consistent representative stochastic trajectory
\begin{equation}
    \dot{x} = x^{\alpha}-A_sx^{\beta+\gamma}-x^{\beta}\big( \mu N \langle x^{\gamma}\rangle + \sigma \sqrt{N} \eta\big) \, ,
\label{eq: dmft}
\end{equation}
where $A_s$ is a random variable with the statistics of $A_{ii}$ and $\eta$ is a Gaussian noise with zero mean and correlation $\langle x^{\gamma}(t)x^{\gamma}(s)\rangle$.

\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{figs/alpha-beta.pdf}
    \caption{The ``May" and ``anti-May" phases described in the
    case of uniform interactions are robust with respect to random interactions.
    Here we simulated system~\eqref{eq: full abg} for a fixed value of
    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates 
    and evaluated $\sigma_c$, defined here as
    the value above which full stable coexistence is less probable than 50\%,
    at varying $\alpha$ and $\beta$.
    No amount of heterogeneity is allowed in the lower triangle because we have $\mu=\mu_s$, while an increasing value of $\sigma_c$
    is found as $\beta$ becomes bigger than $\alpha$.
    We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which instability is triggered also in the upper triangle,
    does not imply that increasing $N$ will destabilize the system, because $\mu N$ would also change.
    }
    \label{fig: alpha-beta}
\end{figure}

@@

Let us focus on the choice 
$\alpha=1$, $\beta=3/2$ and $\gamma=1$, with
$A_s=0$. 
In this case the first and second moments formally diverge.
This is not truly problematic because we are dealing with large, but \emph{finite} populations.
Following~\cite{Cui2020,Hatton2023} and assuming
$\sigma \sqrt{N\langle x_*^2\rangle}\ll \mu N \langle x_* \rangle$,
we can expand the  
solution \eqref{eq: cavity solution} to first order in 
$\sigma \sqrt{N\langle x_*^2\rangle}$, and 
approximate \eqref{eq: dist general} with
a Gaussian distribution with moments $\langle x_*\rangle=(\mu N)^{-2/3}$ and $\langle x_*^2\rangle=(\mu N)^{-4/3}/(1-4(\mu N)^{-2}\sigma^2N)$.

>>>>

\emph{Random interactions: equilibrium distribution.---}Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=-x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
(Any coefficients can be reabsorbed in the statistics of $A$ and by a rescaling of time. 
We factor out a minus sign in order to connect more directly with competition models.)

\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10-diversity.pdf}
    \caption{For a system in the ``anti-May" phase, an increase in $N$,
    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane, will be stabilizing rather than destabilizing.
    Here we show results from the numerical resolution of the dynamical system~\eqref{dynamics}. Stability is defined as full stable coexistence. The red line is computed obtained using DMFT and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
    Parameters values are $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=50$ and $10$ replicates for the simulations.}
    \label{fig: stability line + sims}
\end{figure}

Following e.g. Ref.~\cite{Roy2019}, we can derive from the evolution equation~\eqref{dynamics} a dynamical mean field theory (DMFT) 
which describes the ensemble of $N\gg 1$ variables
by means of a single, self-consistent representative stochastic trajectory
\begin{equation}
    \dot{x} = x^{\alpha}-A_sx^{\beta+\gamma}-x^{\beta}\big( \mu N \langle x^{\gamma}\rangle + \sigma \sqrt{N} \eta\big) \, ,
\label{eq: dmft}
\end{equation}
where $A_s$ is a random variable with the statistics of $A_{ii}$ and $\eta$ is a Gaussian noise with zero mean and correlation $\langle x^{\gamma}(t)x^{\gamma}(s)\rangle$.

\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{figs/alpha-beta.pdf}
    \caption{The ``May" and ``anti-May" phases described in the
    case of uniform interactions are robust with respect to random interactions.
    Here we simulated system~\eqref{dynamics} in the power law case for a fixed value of
    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates 
    and evaluated $\sigma_c$, defined here as
    the value above which full stable coexistence is less probable than 50\%,
    at varying $\alpha$ and $\beta$.
    No amount of heterogeneity is allowed in the lower triangle because we have $\mu=\mu_s$, while an increasing value of $\sigma_c$
    is found as $\beta$ becomes bigger than $\alpha$.
    We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which instability is triggered also in the upper triangle,
    does not imply that increasing $N$ will destabilize the system, because $\mu N$ would also change.
    }
    \label{fig: alpha-beta}
\end{figure}

@@

Let us focus on the choice 
$\alpha=1$, $\beta=3/2$ and $\gamma=1$, with
$A_s=0$. 
In this case the first and second moments formally diverge.
This is not truly problematic because we are dealing with large, but \emph{finite} populations.
Following~\cite{Cui2020,Hatton2023} and assuming
$\sigma \sqrt{N\langle x_*^2\rangle}\ll \mu N \langle x_* \rangle$,
we can expand the  
solution~\eqref{eq: cavity solution} to first order in 
$\sigma \sqrt{N\langle x_*^2\rangle}$, and 
approximate~\eqref{eq: dist general} with
a Gaussian distribution with moments $\langle x_*\rangle=(\mu N)^{-2/3}$ and $\langle x_*^2\rangle=(\mu N)^{-4/3}/(1-4(\mu N)^{-2}\sigma^2N)$.

### 2023-05-02 papers/model.tex
update draft

Consider the dynamical system in $N$ variables
\begin{equation}\label{dynamics}
    \dot{x}_i = f(x_i) + \sum_{j}A_{ij}g(x_i)h(x_j) \, .
\end{equation}
Here $f(x_i)$ represents the self-interaction of a population $i$, while $g(x_i)$ and $h(x_j)$ capture the cross-interaction of $i$ with other populations. That interaction is weighted by a coefficient $A_{ij}$, such that $A_{ij} > 0$ (resp.
$A_{ji} < 0$) implies a positive (resp.
negative) effect of $j$ on the growth of $i$. 
We all assume all of $f$, $g$ and $h$ are increasing functions, and refer to $f$ as the ``production function'' and $g$ as the ``response function''.

>>>>

\emph{Model.---}Consider the dynamical system in $N$ variables
\begin{equation}\label{dynamics}
    \dot{x}_i = f(x_i) + \sum_{j}A_{ij}g(x_i)h(x_j) \, .
\end{equation}
Here $f(x_i)$ represents the self-interaction of a population $i$, while $g(x_i)$ and $h(x_j)$ capture the cross-interaction of $i$ with other populations. That interaction is weighted by a coefficient $A_{ij}$, such that $A_{ij} > 0$ (resp.
$A_{ji} < 0$) implies a positive (resp.
negative) effect of $j$ on the growth of $i$. 
We assume all of $f$, $g$ and $h$ are increasing functions, and refer to $f$ as the ``production function'' and $g$ as the ``response function''.

### 2023-05-02 papers/stability.tex
update draft

We now consider the case of random interactions.
Speficially, we assume that interaction coefficients $A_{ij}$ are drawn independently from a distribution with mean $\mu$ and standard deviation $\sigma$;
We assume the diagonal elements $A_{ii}$ have a mean value $\mu_s$ (possibly different from $\mu$) but the same standard deviation $\sigma$.

@@

To use this result, we decompose the interaction matrix as $A = \mu\mathbf{1} + (\mu_s-\mu)I + S$,
with $\mathbf{1}$ the matrix with all entries equal to $1$ and $S$ a random matrix as defined above.
Up to the rank-one perturbation $\mu\mathbf{1}$ which does not affect stability properties \cite{Stone2018}, we can write the Jacobian \eqref{eq: jac off-diag} as $J = M + LSR$ with diagonal matrices
\begin{equation}
    M_i = g(x_i^*)h'(x_i^*)\psi(x_i^*),\quad L = g(x_i^*), \quad R = h'(x_i^*),
\end{equation}
where $\psi(x)$ is the function defined in \eqref{eq: psi}.

\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10-diversity.pdf}
    \caption{For a system in the ``anti-May" phase, an increase in $N$,
    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane, will be stabilizing rather than destabilizing.
    Here we show results from the numerical resolution of the dynamical system \eqref{dynamics}. Stability is defined as full stable coexistence. The red line is computed obtained using DMFT and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
    Parameters values are $\alpha=1$, $\beta=1.5$,
    $\gamma=1$, $N=50$ and $10$ replicates for the simulations.}
    \label{fig: stability line + sims}
\end{figure}

Stability of the equilibrium requires that $\mathcal{D}$ be entirely contained in the left half-plane (all eigenvalues have negative real part). For this, it is no longer sufficient that $\psi(x_i^*) > 0$ for each $i$: we must also have that $0\notin \mathcal{D}$, hence from \eqref{eq: domain}:
\begin{equation}
    \sum_{i=1}^N \Big(\frac{\sigma}{\psi(x_i^*)}\Big)^{2}
    < 1.
    \label{eq: random-stability}
\end{equation}
In the GLV model, $\psi(x) = \mu-\mu_s$ (independently of $x$) and we recover the classical condition $\sigma\sqrt{N} < (\mu_s-\mu)$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}. With more general power laws, the dependence on equilibrium values $x_i^*$ does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition \eqref{eq: random-stability}.

>>>>

\emph{Random interactions: stability condition.---}We now consider the case of random interactions.
Speficially, we assume that interaction coefficients $A_{ij}$ are drawn independently from a distribution with mean $\mu$ and standard deviation $\sigma$;
We assume the diagonal elements $A_{ii}$ have a mean value $\mu_s$ (possibly different from $\mu$) but the same standard deviation $\sigma$.

@@

To use this result, we decompose the interaction matrix as $A = \mu\mathbf{1} + (\mu_s-\mu)I + S$,
with $\mathbf{1}$ the matrix with all entries equal to $1$ and $S$ a random matrix as defined above.
Up to the rank-one perturbation $\mu\mathbf{1}$ which does not affect stability properties \cite{Stone2018}, we can write the Jacobian~\eqref{eq: jac off-diag} as $J = M + LSR$ with diagonal matrices
\begin{equation}
    M_i = g(x_i^*)h'(x_i^*)\psi(x_i^*),\quad L = g(x_i^*), \quad R = h'(x_i^*),
\end{equation}
where $\psi(x)$ is the function defined in~\eqref{eq: psi}.

\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{figs/cavity.pdf}
    \caption{The equilibrium distribution with power-law self- and cross-interactions is accurately reproduced by 
    Eq.~\eqref{eq: dist general} and its Gaussian approximation.
    For simulations (in green) we used $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=50$, $\mu=\mu_s=10^{-2}$, $\sigma=5\cdot 10^{-3}$.}
    \label{fig: cavity sol.}
\end{figure}

Stability of the equilibrium requires that $\mathcal{D}$ be entirely contained in the left half-plane (all eigenvalues have negative real part). For this, it is no longer sufficient that $\psi(x_i^*) > 0$ for each $i$: we must also have that $0\notin \mathcal{D}$, hence from~\eqref{eq: domain}:
\begin{equation}
    \sum_{i=1}^N \Big(\frac{\sigma}{\psi(x_i^*)}\Big)^{2}
    < 1.
    \label{eq: random-stability}
\end{equation}
In the GLV model, $\psi(x) = \mu-\mu_s$ (independently of $x$) and we recover the classical condition $\sigma\sqrt{N} < (\mu_s-\mu)$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}. In general, the dependence on equilibrium values $x_i^*$ (which, in turn, can depend on $N$) does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition~\eqref{eq: random-stability}.
